UniScatter: Unifying\\Backscatter Communications, Symbiotic Radio and Reconfigurable Intelligent Surface

### 2022-08-20 main.tex
:pencil: Update threshold design

	\emph{Notations:}
	Italic, bold lower-case, and bold upper-case letters denote scalars, vectors and matrices, respectively.
	$\boldsymbol{0}$ and $\boldsymbol{1}$ denote zero and one array of appropriate size, respectively. $\mathbb{I}^{x \times y}$, $\mathbb{R}_+^{x \times y}$, and $\mathbb{C}^{x \times y}$ denote the unit, real nonnegative, and complex spaces of dimension $x \times y$, respectively.
	$j$ denotes the imaginary unit.
	% $\arg(\cdot)$, $\mathrm{rank}(\cdot)$, $\mathrm{tr}(\cdot)$, $\mathrm{diag}(\cdot)$ and $\mathrm{diag}^{-1}(\cdot)$ denote the argument, rank, trace, a square matrix with input vector on main diagonal, and a vector retrieving main diagonal of input square matrix, respectively.
	$\mathrm{diag}(\cdot)$ returns a square matrix with the input vector on its main diagonal and zeros elsewhere.
	$(\cdot)^*$, $(\cdot)^T$, $(\cdot)^H$, $\lvert{\cdot}\rvert$, and $\lVert{\cdot}\rVert$ denote the conjugate, transpose, conjugate transpose, absolute value, and Euclidean norm operators, respectively.
	$(\cdot)^{(r)}$ and $(\cdot)^{\star}$ denote the $r$-th iterated and terminal solutions, respectively.
	The distribution of a \gls{cscg} random variable with zero mean and variance $\sigma^2$ is denoted by $\mathcal{CN}(0,\sigma^2)$, and $\sim$ means ``distributed as''.
\end{section}

@@

		For the ease of notations, we define a general bijective function transforming backscatter index tuple $m_{\mathcal{K}}$ to integer $l \in \mathcal{L} \triangleq \{1,\ldots,L \triangleq M^K\}$, and use both notation interchangeably in the following context.
		As such, the convex decision region of backscatter index tuple $m_{\mathcal{K}}$ or equivalent letter $l$ can be written as

@@

			\mathcal{R}_{m_{\mathcal{K}}} \triangleq [t_{l-1},t_l), \quad t_{l-1} \le t_l.

@@

\begin{section}{Rate-Region Characterization}
	To characterize the achievable primary-total-backscatter rate region of the proposed UniScatter system, we aim to maximize the weighted sum mutual information with respect to input distribution $\{\boldsymbol{p}_k\}_{k \in \mathcal{K}}$, active beamforming $\boldsymbol{w}$, and decision thresholds $\boldsymbol{t}$ as
	\begin{maxi!}
		{\scriptstyle{\{\boldsymbol{p}_k\}_{k \in \mathcal{K}},\boldsymbol{w},\boldsymbol{t}}}{I(x_{\mathcal{K}})}{\label{op:weighted_sum_rate}}{\label{ob:weighted_sum_rate}}
		\addConstraint{\boldsymbol{1}^T \boldsymbol{p}_k}{=1,}{\quad \forall k}{\label{co:sum_probability}}
		\addConstraint{\boldsymbol{p}_k}{\ge \boldsymbol{0},}{\quad \forall k}{\label{co:nonnegative_probability}}
		\addConstraint{\lVert \boldsymbol{w} \rVert^2}{\le P}{\label{co:transmit_power}}
		\addConstraint{t_{l-1}}{\le t_l,}{\quad \forall l.}{\label{co:decision_threshold}}
	\end{maxi!}
	Since problem \eqref{op:weighted_sum_rate} is highly non-convex, we propose a \gls{bcd} algorithm that iteratively updates $\{\boldsymbol{p}_k\}_{k \in \mathcal{K}}$, $\boldsymbol{w}$ and $\boldsymbol{t}$ until convergence.

@@

	\begin{subsection}{Decision Threshold}
		For any given $\{\boldsymbol{p}_k\}_{k \in \mathcal{K}}$ and $\boldsymbol{w}$, problem \eqref{op:weighted_sum_rate} reduces to
		\begin{maxi!}
			{\scriptstyle{\boldsymbol{t}}}{I(x_{\mathcal{K}})}{\label{op:decision_threshold}}{\label{ob:decision_threshold}}
			\addConstraint{\eqref{co:decision_threshold},}
		\end{maxi!}
		which is still non-convex because variable $\boldsymbol{t}$ appear on the limits of integration \eqref{eq:dmtmac}.
		To tackle this, we reveal backscatter decision design has no impact on primary achievable rate, and regard the dispersed nodes with given input distribution as an equivalent augmented source in terms of total rate.
		% We then constrain the feasible domain of problem \eqref{op:decision_threshold} from continuous space $\mathbb{R}_+^{L+1}$ to finite candidate set (i.e., fine-grained discrete energy levels) $\mathcal{T}^{L+1}$, and reuse existing rate-optimal quantization schemes for \gls{dmc} to solve problem \eqref{op:decision_threshold}.

		\begin{remark}
			Upon successful backscatter decoding, UniScatter receiver can always re-encode node messages to recover their reflection patterns at each primary block.
			Therefore, backscatter decision design has no impact on the primary achievable rate, and any thresholding scheme maximizing the total backscatter mutual information \eqref{eq:backscatter_mutual_information} is also optimal for problem \eqref{op:decision_threshold}.
			\label{re:backscatter_detection}
		\end{remark}

		\begin{remark}
			In terms of total backscatter rate, the dispersed nodes with given input distribution can be viewed as an equivalent source with augmented alphabet of backscatter symbol tuples $x_{m_{\mathcal{K}}}$.
			As such, the \gls{dmtmac} \eqref{eq:dmtmac} is essentially a \gls{dmtc}, and problem \eqref{op:decision_threshold} reduces to rate-optimal quantization design for discrete-input continuous-output memoryless channel.
		\end{remark}

		Next, we constrain the feasible domain of problem \eqref{op:decision_threshold} from continuous space $\mathbb{R}_+^{L+1}$ to finite candidate set (i.e., fine-grained discrete energy levels) $\mathcal{T}^{L+1}$, and reuse existing rate-optimal quantization schemes for \gls{dmc} to solve problem \eqref{op:decision_threshold}.

		Remark \ref{re:backscatter_detection} suggests any $\boldsymbol{t}$ maximizes total backscatter mutual information $I_{\mathrm{B}}(x_{\mathcal{K}};\hat{x}_{\mathcal{K}})$ is also optimal for problem \eqref{op:decision_threshold}.

		where $t_0^\star=0$ and $t_L^\star=\infty$ for energy-based backscatter detection.

		\begin{remark}
			% With input distribution of all nodes, we can formulate an equivalent information source with augmented alphabet of tag input combination.
			% In terms of total backscatter rate, we can formulate an equivalent information source with augmented alphabet of tag input combination.
			% When input distribution of all nodes are available, we can formulate an equivalent information source in terms of total backscatter rate.
			% When input distribution of all nodes are available, we can think an equivalent information source with augmented alphabet of state tuple, in terms of total backscatter rate.
			% In terms of total backscatter rate, we can regard all nodes as an equivalent information source with augmented alphabet of node state tuple $x_{m_{\mathcal{K}}}$.
			In terms of total backscatter rate, the nodes can be regarded as an equivalent source with augmented alphabet of symbol tuple $x_{m_{\mathcal{K}}}$, and the \gls{dmtmac} \eqref{eq:dmtmac} essentially reduces to a \gls{dmtc}.
			% As such, the \gls{dmtmac} \eqref{eq:dmtmac} is essentially a \gls{dmtc}, and decision threshold design can be simplified accordingly.
		\end{remark}

		Finally, we can employ existing thresholding design for \gls{dmtc} to solve problem \eqref{op:decision_threshold}.
		For example, \cite{He2021} first discretized the continuous energy $z$ into numerous output bins, then grouped adjacent bins to maximize mutual information using \gls{dp} accelerated by \gls{smawk} algorithm.
		In \cite{Nguyen2020a}, the authors first proved the optimality condition for any three neighbor thresholds, then fix $t_0$, traverse $t_1$, and sequentially optimizes the others by bisection.
		Both will be compared with \gls{ml} decision \cite{Qian2019}
		\begin{equation}
			t_{l}^{\mathrm{\gls{ml}}} = N \frac{\sigma_{l-1}^2 \sigma_{l}^2}{\sigma_{l-1}^2 - \sigma_{l}^2} \log \frac{\sigma_{l-1}^2}{\sigma_{l}^2}, \quad l \in \mathcal{L} \setminus \{0,L\},
			\label{eq:detection_threshold_maximum_likelihood}
		\end{equation}
		which is generally suboptimal for problem \eqref{op:decision_threshold} except for equiprobable inputs at all nodes.

		% \begin{remark}
		% 	Although the \gls{ml} decision threshold is optimal in terms of the likelihood function, it is not necessarily capacity-achieving, although the performance gap can be negligible in the single-tag \gls{bibo} case.
		% \end{remark}

		% Interestingly, the optimal threshold design to maximize the mutual information for a general \gls{dmtc} with a fixed number of output letters remains an open issue.
		% The reason is that each decision region may contain more than one disjoint partitions (i.e., non-convex) and the number of thresholds are unknown.
		% Fortunately, for the proposed energy detection, we proved that the \gls{dmtc} capacity can be achieved using only convex decision regions.
		% This conclusion is summarized below.

		% \begin{proposition}
		% 	For a discrete-input continuous-output channel in Gamma distribution \eqref{eq:energy_distribution}, if the \gls{dmtc} is constructed for detection (i.e., same input/output alphabet) and $L$ input letters are with non-zero probability, then it is possible to achieve the \gls{dmtc} capacity by $L$ non-empty convex decision regions defined by $L+1$ distinct decision thresholds.
		% 	\label{pr:threshold}
		% \end{proposition}

		% \begin{proof}
		% 	Please refer to Appendix \ref{ap:threshold}.
		% 	\label{pf:threshold}
		% \end{proof}

		% Once the optimal number of decision threshold is determined, we can first discretize the output energy level into numerous bins, then obtain the optimal decision regions that maximizes the total backscatter rate by \gls{dp} accelerated by \gls{smawk} algorithm \cite{He2021}.
	\end{subsection}
\end{section}

\begin{section}{Simulation Results}
	% TODO add distribution to initializer to smooth curves?
	% TODO compare PGD and primary-MRT beamformers
	In this section, we provide numerical results to evaluate the proposed input, beamforming and decision design over a single-user multi-node UniScatter network.
	We assume the distance between \gls{ap} and user is \qty{10}{\meter}, and $K=2$ UniScatter nodes are uniformly dropped within a disk centered at the user of radius \qty{1}{\meter}.
	% The carrier frequency is $f=\qty{200}{\MHz}$, and we consider \gls{iid} Ricean fading for all channels.
	The carrier frequency is $f=\qty{200}{\MHz}$, and we consider \gls{iid} Ricean fading between all terminals.
	For direct, forward and backward links, we set the path loss exponents to \num{2.6}, \num{2.4} and \num{2}, and the Ricean factor to \num{5}, \num{5} and \num{10}, respectively.
	% For direct, forward and backward links, we choose path loss exponents \num{2.6}, \num{2.4} and \num{2}, and Ricean factor \num{5}, \num{5} and \num{10}, respectively.
	% and the path loss exponents for direct, forward and backward links are \num{2.6}, \num{2.4} and \num{2}, respectively.
	% We consider \gls{iid} Ricean fading for all channels with Ricean factor set to \num{5}, \num{5} and \num{10}
	The \gls{ap} has $Q=4$ antennas with maximum average transmit power $P=\qty{36}{\dBm}$.
	All nodes have amplitude scattering ratio $\alpha=0.5$, symbol period ratio $N=20$, and perform \gls{qam} with $M=4$ input states.
	% and we consider amplitude scattering ratio $\alpha=0.5$ and symbol period ratio $N=20$.
	% The user has a receive antenna gain of \qty{3}{\dBi} and average noise power
	The user is with average noise power $\sigma_v^2=\qty{70}{\dBm}$ and receive antenna gain \qty{3}{\dBi}.
	All rate regions are averaged over at least \num{1000} instances, where ``PSP'' and ``BSP'' means primary and backscatter symbol periods, respectively.
	We choose decision design by \gls{smawk} \cite{He2021} as reference, and select discretize boundaries uniformly over the \qty{95}{\percent} confidence region of edge hypotheses.
	The parameters remain fixed unless otherwise specified.

@@

		Fig. \subref*{fi:region_coverage} shows the tradeoff between coverage disk radius $r$ and achievable rate region.
		When nodes are far from the user, both primary and backscatter rates decrease due to the product pass loss of forward and backward channels.
		Under the assumption of ideal backscatter decoding and re-encoding, Fig. \subref*{fi:region_duration} suggests using lower symbol period ratio $N$ can boost total backscatter rate per primary symbol.
		However, it requires more frequent detection and re-encoding at the user to maintain the primary rate.
		When $N$ becomes sufficiently large, total backscatter rate approaches \num{0} and UniScatter nodes boil down to conventional \gls{ris} elements with fixed reflection pattern during whole channel block.
		In Fig. \subref*{fi:region_frequency}, we observe that passive UniScatter nodes achieve higher backscatter rate at lower carrier frequency because of preferable propagation loss.
		Finally, \subref*{fi:region_noise} prove the performance of energy detection is robust for a wide range of noise power.
	\end{subsection}
\end{section}

>>>>

	\emph{Notations:}
	Italic, bold lower-case, and bold upper-case letters denote scalars, vectors and matrices, respectively.
	$\boldsymbol{0}$ and $\boldsymbol{1}$ denote zero and one array of appropriate size, respectively. $\mathbb{I}^{x \times y}$, $\mathbb{R}_+^{x \times y}$, and $\mathbb{C}^{x \times y}$ denote the unit, real nonnegative, and complex spaces of dimension $x \times y$, respectively.
	$j$ denotes the imaginary unit.
	% $\arg(\cdot)$, $\mathrm{rank}(\cdot)$, $\mathrm{tr}(\cdot)$, $\mathrm{diag}(\cdot)$ and $\mathrm{diag}^{-1}(\cdot)$ denote the argument, rank, trace, a square matrix with input vector on main diagonal, and a vector retrieving main diagonal of input square matrix, respectively.
	$\mathrm{diag}(\cdot)$ returns a square matrix with the input vector on its main diagonal and zeros elsewhere.
	$\mathrm{card}(\cdot)$ returns the cardinality of a set.
	$(\cdot)^*$, $(\cdot)^T$, $(\cdot)^H$, $\lvert{\cdot}\rvert$, and $\lVert{\cdot}\rVert$ denote the conjugate, transpose, conjugate transpose, absolute value, and Euclidean norm operators, respectively.
	$(\cdot)^{(r)}$ and $(\cdot)^{\star}$ denote the $r$-th iterated and terminal solutions, respectively.
	The distribution of a \gls{cscg} random variable with zero mean and variance $\sigma^2$ is denoted by $\mathcal{CN}(0,\sigma^2)$, and $\sim$ means ``distributed as''.
\end{section}

@@

		For the ease of notations, we define a general bijective function from backscatter index tuple $m_{\mathcal{K}}$ to integer $l \in \mathcal{L} \triangleq \{1,\ldots,L \triangleq M^K\}$, where $\{\sigma_l^2\}_{l \in \mathcal{L}}$ are sorted in an ascending order.
		Both notations are used interchangeably in the following context.
		As such, the convex decision region of backscatter tuple $l$ can be written as

@@

			\mathcal{R}_{l} \triangleq [t_{l-1},t_l), \quad t_{l-1} \le t_l.

@@

\begin{section}{Rate-Region Characterization}
	To characterize the achievable primary-total-backscatter rate region of the proposed UniScatter system, we aim to maximize the weighted sum mutual information with respect to input distribution $\{\boldsymbol{p}_k\}_{k \in \mathcal{K}}$, active beamforming $\boldsymbol{w}$, and decision thresholds $\boldsymbol{t}$ as
	\begin{maxi!}
		{\scriptstyle{\{\boldsymbol{p}_k\}_{k \in \mathcal{K}},\boldsymbol{w},\boldsymbol{t}}}{I(x_{\mathcal{K}})}{\label{op:weighted_sum_rate}}{\label{ob:weighted_sum_rate}}
		\addConstraint{\boldsymbol{1}^T \boldsymbol{p}_k}{=1,}{\quad \forall k}{\label{co:sum_probability}}
		\addConstraint{\boldsymbol{p}_k}{\ge \boldsymbol{0},}{\quad \forall k}{\label{co:nonnegative_probability}}
		\addConstraint{\lVert \boldsymbol{w} \rVert^2}{\le P}{\label{co:transmit_power}}
		% \addConstraint{t_{l-1}}{\le t_l,}{\quad \forall l.}{\label{co:decision_threshold}}
		\addConstraint{\boldsymbol{t}}{\ge \boldsymbol{0}.}{\label{co:decision_threshold}}
	\end{maxi!}
	Since problem \eqref{op:weighted_sum_rate} is highly non-convex, we propose a \gls{bcd} algorithm that iteratively updates $\{\boldsymbol{p}_k\}_{k \in \mathcal{K}}$, $\boldsymbol{w}$ and $\boldsymbol{t}$ until convergence.

@@

	\begin{subsection}{Decision Threshold}
		For any given $\{\boldsymbol{p}_k\}_{k \in \mathcal{K}}$ and $\boldsymbol{w}$, problem \eqref{op:weighted_sum_rate} reduces to
		\begin{maxi!}
			{\scriptstyle{\boldsymbol{t}}}{I(x_{\mathcal{K}})}{\label{op:decision_threshold}}{\label{ob:decision_threshold}}
			\addConstraint{\eqref{co:decision_threshold},}
		\end{maxi!}
		which is still non-convex because variable $\boldsymbol{t}$ appears on the limits of integration \eqref{eq:dmtmac}.
		Fortunately, we can further simplify problem \eqref{op:decision_threshold} as a point-to-point rate-optimal quantization design for a discrete-input continuous-output memoryless channel, thanks to Remark \ref{re:backscatter_decision} and \ref{re:augmented_source}.

		\begin{remark}
			Upon successful backscatter decoding, UniScatter receiver can always re-encode node messages to recover their reflection patterns at each primary block.
			Therefore, backscatter decision design has no impact on the primary achievable rate, and any thresholding scheme maximizing the total backscatter mutual information \eqref{eq:backscatter_mutual_information} is also optimal for problem \eqref{op:decision_threshold}.
			\label{re:backscatter_decision}
		\end{remark}

		\begin{remark}
			In terms of total backscatter rate, the dispersed nodes with given input distribution can be viewed as an equivalent source with augmented alphabet of backscatter symbol tuples $\{x_{m_{\mathcal{K}}}\}$.
			As such, the \gls{dmtmac} \eqref{eq:dmtmac} is essentially a \gls{dmtc}, and problem \eqref{op:decision_threshold} reduces to the rate-optimal quantization design for a discrete-input continuous-output memoryless channel.
			\label{re:augmented_source}
		\end{remark}

		Next, we constrain the feasible domain of problem \eqref{op:decision_threshold} from continuous space $\mathbb{R}_+^{L+1}$ to finite candidate set (i.e., fine-grained discrete energy levels) $\mathcal{T}^{L+1}$.
		\begin{figure}[!t]
			\centering
			\resizebox{0.9\columnwidth}{!}{
				\input{assets/illustration/discrete_outputs.tex}
			}
			\caption{
				The decision thresholds are selected from fine-grained discrete energy levels instead of continuous space, and each decision region consists of at least one neighbor energy bins.
			}
			\label{fi:discrete_outputs}
		\end{figure}
		As shown in Fig. \ref{fi:discrete_outputs}, by introducing the extra analog-to-digital conversion step, we can obtain decision regions by grouping adjacent high-resolution energy bins.
		Thus, problem \eqref{op:decision_threshold} is recast as
		\begin{maxi}
			{\scriptstyle{\boldsymbol{t} \in \mathcal{T}^{L+1}}}{I_{\mathrm{B}}(x_{\mathcal{K}};\hat{x}_{\mathcal{K}}),}{\label{op:decision_threshold_recast}}{}
		\end{maxi}
		which can be solved by existing rate-optimal \gls{dmc} quantizer designs.
		For global optimal solutions, \cite{He2021} proposed a \gls{dp} method accelerated by the \gls{smawk} algorithm with computational complexity $\mathcal{O}\bigl(L^2(\mathrm{card}(\mathcal{T})-L)\bigr)$, and \cite{Nguyen2020a} presented a traverse-bisection algorithm using the optimality condition of three neighbor thresholds with complexity $\mathcal{O}\bigl(\mathrm{card}(\mathcal{T})L\log(\mathrm{card}(\mathcal{T})L)\bigr)$.
		In Section \ref{se:simulation_results}, both schemes will be compared with the \gls{ml} thresholding scheme \cite{Qian2019}
		\begin{equation}
			t_{l}^{\mathrm{\gls{ml}}} = N \frac{\sigma_{l-1}^2 \sigma_{l}^2}{\sigma_{l-1}^2 - \sigma_{l}^2} \log \frac{\sigma_{l-1}^2}{\sigma_{l}^2}, \quad l \in \mathcal{L} \setminus \{L\},
			\label{eq:detection_threshold_maximum_likelihood}
		\end{equation}
		which is generally suboptimal for problem \eqref{op:decision_threshold} except when all nodes are with equiprobable inputs.
	\end{subsection}
\end{section}

\begin{section}{Simulation Results}
	% TODO design $\mathcal{T}$
	% TODO add distribution to initializer to smooth curves?
	% TODO compare PGD and primary-MRT beamformers
	In this section, we provide numerical results to evaluate the proposed input, beamforming and decision design over a single-user multi-node UniScatter network.
	We assume the distance between \gls{ap} and user is \qty{10}{\meter}, and $K=2$ UniScatter nodes are uniformly dropped within a disk centered at the user of radius \qty{1}{\meter}.
	% The carrier frequency is $f=\qty{200}{\MHz}$, and we consider \gls{iid} Ricean fading for all channels.
	The carrier frequency is $f=\qty{200}{\MHz}$, and we consider \gls{iid} Ricean fading between all terminals.
	For direct, forward and backward links, we set the path loss exponents to \num{2.6}, \num{2.4} and \num{2}, and the Ricean factor to \num{5}, \num{5} and \num{10}, respectively.
	% For direct, forward and backward links, we choose path loss exponents \num{2.6}, \num{2.4} and \num{2}, and Ricean factor \num{5}, \num{5} and \num{10}, respectively.
	% and the path loss exponents for direct, forward and backward links are \num{2.6}, \num{2.4} and \num{2}, respectively.
	% We consider \gls{iid} Ricean fading for all channels with Ricean factor set to \num{5}, \num{5} and \num{10}
	The \gls{ap} has $Q=4$ antennas with maximum average transmit power $P=\qty{36}{\dBm}$.
	All nodes have amplitude scattering ratio $\alpha=0.5$, symbol period ratio $N=20$, and perform \gls{qam} with $M=4$ input states.
	% and we consider amplitude scattering ratio $\alpha=0.5$ and symbol period ratio $N=20$.
	% The user has a receive antenna gain of \qty{3}{\dBi} and average noise power
	The user is with average noise power $\sigma_v^2=\qty{70}{\dBm}$ and receive antenna gain \qty{3}{\dBi}.
	All rate regions are averaged over at least \num{1000} instances, where ``PSP'' and ``BSP'' means primary and backscatter symbol periods, respectively.
	We choose decision design by \gls{smawk} \cite{He2021} as reference, and select discretize boundaries uniformly over the \qty{95}{\percent} confidence region of edge hypotheses.
	The parameters remain fixed unless otherwise specified.

@@

		Fig. \subref*{fi:region_coverage} shows the tradeoff between coverage disk radius $r$ and achievable rate region.
		When nodes are far from the user, both primary and backscatter rates decrease due to the product pass loss of forward and backward channels.
		Under the assumption of ideal backscatter decoding and re-encoding, Fig. \subref*{fi:region_duration} suggests using lower symbol period ratio $N$ can boost total backscatter rate per primary symbol.
		However, it requires more frequent detection and re-encoding at the user to maintain the primary rate.
		When $N$ becomes sufficiently large, total backscatter rate approaches \num{0} and UniScatter nodes boil down to conventional \gls{ris} elements with fixed reflection pattern during whole channel block.
		In Fig. \subref*{fi:region_frequency}, we observe that passive UniScatter nodes achieve higher backscatter rate at lower carrier frequency because of preferable propagation loss.
		Finally, \subref*{fi:region_noise} prove the performance of energy detection is robust for a wide range of noise power.
	\end{subsection}
	\label{se:simulation_results}
\end{section}
